---
title: "Visualizing the NTL LTER Data"
author: "Maxwell Pepperdine"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Load packages

```{r}
rm(list=ls())
```

```{r}
library(tidyverse)
library(DBI)
library(duckdb)
library(dbplyr)
library(here)
```

## Background

This script will initiate a connection with the `ntl_lter_database.db` to create a query that answers the following questions:

- What are the average ice duration and meteorological data for the `ME` lake versus the `MO` lakes? (**primary question**)
- What's the relationship between yearly ice duration in the Madison Lakes, WI area and the following meteorological data: `avg max air temp`, `avg min air temp`, `avg air temp`, `precip`, and `snow`? (**secondary question**)

#### Data

See the **Data Access** section of the `README.md` file for information on sourcing the datasets used in this analysis. The data queried and visualized here is from the `ice_duration` and `yearly_iceszn_meteor_records` tables in the `database/ntl_lter_database.db` database.

These data were cleaned in R in the `NTL-LTER_data_cleaning.qmd` file, and then ingested into a DuckDB database This database is stored in the `database/ntl_lter_database.db` file.

## Connect to the NTL LTER database

```{r}
conn <- DBI::dbConnect(duckdb::duckdb(), dbdir = here("database/ntl_lter_database.db"))
```

#### Take a look at our tables

```{r}
DBI::dbListTables(conn)
```

## Make the queries

#### Get each table

```{r}
ice_duration <- tbl(conn, "ice_duration")
meteor_records <- tbl(conn, "yearly_iceszn_meteor_records")
```

```{r}
# convert the tables to dfs to view the data easier
ice_duration_df <- ice_duration %>%
  collect()
meteor_records_df <- meteor_records %>%
  collect()
```

#### Make our query with `dbplyr`

###### Compare `ME` and `MO` lakes

This query aims to answer the following question:

-   How do the average ice duration and meteorological data (max air temp adjusted, min air temp adjusted, avg air temp adjusted, precip, snow) vary between the `ME` and `MO` lake?

It will join the `ice_duration` and `yearly_iceszn_meteor_records` tables on the `year` column, and then group the data by `lakeid`. It will then calculate the average ice duration and meteorological data for each lake, returning a table with this information. 

```{r}
# generate a query to answer our question
ntl_joined <- ice_duration %>% 
  left_join(meteor_records, 
                        by = c("year" = "year")) %>% 
  group_by(lakeid) %>%
  summarise(avg_ice_duration = mean(duration, na.rm = TRUE),
            avg_max_air_temp = mean(max_air_temp_adjusted, na.rm = TRUE),
            avg_min_air_temp = mean(min_air_temp_adjusted, na.rm = TRUE),
            avg_air_temp = mean(avg_air_temp_adjusted, na.rm = TRUE),
            avg_precip_mm = mean(precip_raw_mm, na.rm = TRUE),
            avg_snow_cm = mean(snow_raw_cm, na.rm = TRUE)) %>% 
  collect()
```

#### Tidy the data for plotting 

```{r}
# pivot the data to tidy it for ggplot
ntl_joined_long <- ntl_joined %>%
  pivot_longer(cols = 2:7,
               names_to = "meteor_variable",
               values_to = "value")
```

## Make some visualizations!

#### Average ice duration and meteorological data for each lake

This visualization will show the average ice duration and meteorological data for the `ME` and `MO` lakes. It is associated with our primary question as stated above in the `Background` section of this script. 

```{r}
# reorder to put the ice duration first 
ntl_joined_long <- ntl_joined_long %>%
  mutate(meteor_variable = factor(meteor_variable, 
                                  levels = c("avg_ice_duration", "avg_max_air_temp", 
                                             "avg_min_air_temp", "avg_air_temp", 
                                             "avg_precip_mm", "avg_snow_cm")))
```

```{r}
# barplot of average ice duration and meteorological data
ggplot(ntl_joined_long, aes(x = meteor_variable, y = value, 
                            fill = lakeid)) +
  geom_bar(stat = "identity", 
           position = "dodge", 
           col = "black", 
           linewidth = 0.5) +
  geom_text(aes(label = round(value, 2)), 
            size = 3, 
            position = position_dodge(width = 0.9), 
            vjust = -0.25) +
  labs(title = "Average ice duration and meteorological data in Lake Mendota (ME) \nand Lake Monona (MO) (1869-2023)",
       x = " ",
       y = "Value", 
       fill = " ") +
  theme_bw() + 
  scale_fill_manual(values = c("ME" = "skyblue4", "MO" = "cadetblue1")) +
  scale_x_discrete(labels=c("avg_ice_duration"="Avg ice duration (days)",
                             "avg_max_air_temp"="Avg max air temp (°C)",
                             "avg_min_air_temp"="Avg min air temp (°C)",
                             "avg_air_temp"="Avg air temp (°C)",
                             "avg_precip_mm"="Precip (mm)",
                             "avg_snow_cm"="Snow (cm)")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Comparing the relationship between ice duration and the meteorological data

Additionally, we can explore the relationship between ice duration and the meteorological data to examine how the average ice duration in the `ME` and `MO` lakes varies with changing temperature, snow, etc. We will only examine average maximum air temperature since this isn't directly related to the assignment, but we could easily do this for all of the meteorological variables.

This figure is associated with our secondary question question as described in the `Background` section of this script. It's not something that we aimed to answer with our main query above, but still an interesting relationship within the data that we can quickly explore!

###### Join the tables

```{r}
# add all meteorological variables to each ice duration record
ntl_joined2 <- left_join(ice_duration, meteor_records, 
                        by = c("year" = "year")) %>%
  collect()
```

###### Filter to only include the years 1950-2023

```{r}
# make another table that only includes the years 1950-2023
ntl_joined_1950_2023 <- ntl_joined2 %>%
  filter(year >= 1950) %>%
  collect()
```

Make a scatterplot of ice duration vs. each meteorological variable.

```{r}
# ice duration vs. avg max air temp

# using all of the data
ggplot(ntl_joined2, aes(x = max_air_temp_adjusted, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Ice duration vs. average max air temprature (1869-2023)",
       x = "Avg max air temp (°C) during ice season",
       y = "Ice duration (days)") +
  facet_wrap(~ lakeid) +
  theme_bw()

# using only the years 1950-2023
ggplot(ntl_joined_1950_2023, aes(x = max_air_temp_adjusted, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Ice duration vs. average max air temprature (1950-2023)",
       x = "Avg max air temp (°C) during ice season",
       y = "Ice duration (days)") +
  facet_wrap(~ lakeid) +
  theme_bw()
```

## Discussion 

#### Primary question

The average ice duration and meteorological data for the `ME` and `MO` lakes are shown in the barplot above. Overall, there's little variation between each of the variables explored. This was to be expected, as Lake Mendota (ME) and Lake Monona (MO) are located in close proximity to each other; however, it was still a fun exercise to explore the data and practice developing queries to answer questions of interest.

#### Secondary question

The scatterplots above show the relationship between ice duration and average maximum air temperature for the `ME` and `MO` lakes. The first plot shows the relationship between ice duration and average maximum air temperature for all of the data (1869-2023), while the second plot shows the same relationship but only for the years 1950-2023. In both plots, there's a clear negative correlation between average maximum air temperature and ice duration (i.e., as the average maximum air temperature increases, ice duration decreases). This makes sense intuitively, as warmer temperatures would lead to shorter ice duration. 










